Plug-and-play dual-tree algorithm runtime analysis

### 2015-08-14 bounds-files/general-runtime-bound.tex
i_t -> I_t, i_n -> I_n.

\begin{equation}
O\left(c_r^4 | R^* | \chi \psi (N + i_t(\mathscr{T}_q) + \theta)\right)
\end{equation}

@@

Next, consider case \textit{(2)}.  In this situation, each query recursion
implies at least one reference recursion before another query recursion.  For
some query node $\mathscr{N}_q$, the exact number of reference recursions before
the children of $\mathscr{N}_q$ are recursed into is bounded above by
$i_n(\mathscr{N}_q) + 1$: if $\mathscr{N}_q$ has imbalance $0$, then it is
exactly one level below its parent, and thus there is only one reference
recursion.  On the other hand, if $\mathscr{N}_q$ is many levels below its
parent, then it is possible that a reference recursion may occur for each level
in between; this is a maximum of $i_n(\mathscr{N}_q) + 1$.

@@

\begin{equation}
\sum_{\mathscr{N}_q \in \mathscr{T}_q} i_n(\mathscr{N}_q) + 1 = i_t(\mathscr{T}_q) +
O(N)
\end{equation}

@@

\begin{equation}
O\left(c_r^4 |R^*| \psi \chi \left(N + i_t(\mathscr{T}_q) + \theta\right)\right)
+ O\left(c_r^4 |R^*| \psi N\right) \sim O\left(c_r^4 |R^*| \psi
\chi \left(N + i_t(\mathscr{T}_q) + \theta\right)\right).
\end{equation}
\end{proof}

@@

\begin{equation}
O\left(c^4 |R^*| \chi \psi \left(N + i_t(\mathscr{T})\right)\right)
\end{equation}

@@

An intuitive understanding of these bounds is best achieved by first considering
the monochromatic case (this case arises, for instance, in all-nearest-neighbor
search).  The linear dependence on $N$ arises from the fact that all query nodes
must be visited.  The dependence on the reference tree, however, is encapsulated
by the term $c^4 |R^*|$, with $|R^*|$ being the maximum size of the reference
set $R$; this value must be derived for each specific problem.  The bad
performance of poorly-behaved datasets with large $c$ (or, in the worst case, $c
\sim N$) is then captured in both of those terms.  Poorly-behaved datasets may
also have a high cover tree imbalance $i_t(\mathscr{T})$; the linear dependence of
runtime on imbalance is thus sensible for well-behaved datasets.

@@

$\nu$ corresponds to the maximum number of reference recursions between a single
query recursion, and $\kappa$ corresponds to the maximum number of query
recursions between a single reference recursion.  The respective proofs that use
these constants then apply them as a worst-case measure for the whole algorithm:
when using $\kappa$, \citet{ram2009} assume that {\it every} reference recursion
may be followed by $\kappa$ query recursions; similarly, \citet{curtin2014dual}
assume that {\it every} query recursion may be followed by $\nu$ reference
recursions.  Here, we have simply used $i_t(\mathscr{T}_q)$ and $\theta$ as an
exact summation of the total extra reference recursions, which gives us a much
tighter bound than $\nu$ or $\kappa$ on the running time of the whole algorithm.

Further, both $\nu$ and $\kappa$ are difficult to empirically calculate and
require an entire run of the dual-tree algorithm.  On the other hand, bounding
$i_t(\mathscr{T}_q)$ (and $\theta$) can be done in one pass of the tree
(assuming the tree is already built).  Thus, not only is our bound tighter when
the cover tree imbalance is sublinear in $N$, it more closely reflects the
actual behavior of dual-tree algorithms, and the constants which it depends upon
are straightforward to calculate.

>>>>

\begin{equation}
O\left(c_r^4 | R^* | \chi \psi (N + I_t(\mathscr{T}_q) + \theta)\right)
\end{equation}

@@

Next, consider case \textit{(2)}.  In this situation, each query recursion
implies at least one reference recursion before another query recursion.  For
some query node $\mathscr{N}_q$, the exact number of reference recursions before
the children of $\mathscr{N}_q$ are recursed into is bounded above by
$I_n(\mathscr{N}_q) + 1$: if $\mathscr{N}_q$ has imbalance $0$, then it is
exactly one level below its parent, and thus there is only one reference
recursion.  On the other hand, if $\mathscr{N}_q$ is many levels below its
parent, then it is possible that a reference recursion may occur for each level
in between; this is a maximum of $I_n(\mathscr{N}_q) + 1$.

@@

\begin{equation}
\sum_{\mathscr{N}_q \in \mathscr{T}_q} I_n(\mathscr{N}_q) + 1 = I_t(\mathscr{T}_q) +
O(N)
\end{equation}

@@

\begin{equation}
O\left(c_r^4 |R^*| \psi \chi \left(N + I_t(\mathscr{T}_q) + \theta\right)\right)
+ O\left(c_r^4 |R^*| \psi N\right) \sim O\left(c_r^4 |R^*| \psi
\chi \left(N + I_t(\mathscr{T}_q) + \theta\right)\right).
\end{equation}
\end{proof}

@@

\begin{equation}
O\left(c^4 |R^*| \chi \psi \left(N + I_t(\mathscr{T})\right)\right)
\end{equation}

@@

An intuitive understanding of these bounds is best achieved by first considering
the monochromatic case (this case arises, for instance, in all-nearest-neighbor
search).  The linear dependence on $N$ arises from the fact that all query nodes
must be visited.  The dependence on the reference tree, however, is encapsulated
by the term $c^4 |R^*|$, with $|R^*|$ being the maximum size of the reference
set $R$; this value must be derived for each specific problem.  The bad
performance of poorly-behaved datasets with large $c$ (or, in the worst case, $c
\sim N$) is then captured in both of those terms.  Poorly-behaved datasets may
also have a high cover tree imbalance $I_t(\mathscr{T})$; the linear dependence of
runtime on imbalance is thus sensible for well-behaved datasets.

@@

$\nu$ corresponds to the maximum number of reference recursions between a single
query recursion, and $\kappa$ corresponds to the maximum number of query
recursions between a single reference recursion.  The respective proofs that use
these constants then apply them as a worst-case measure for the whole algorithm:
when using $\kappa$, \citet{ram2009} assume that {\it every} reference recursion
may be followed by $\kappa$ query recursions; similarly, \citet{curtin2014dual}
assume that {\it every} query recursion may be followed by $\nu$ reference
recursions.  Here, we have simply used $I_t(\mathscr{T}_q)$ and $\theta$ as an
exact summation of the total extra reference recursions, which gives us a much
tighter bound than $\nu$ or $\kappa$ on the running time of the whole algorithm.

Further, both $\nu$ and $\kappa$ are difficult to empirically calculate and
require an entire run of the dual-tree algorithm.  On the other hand, bounding
$I_t(\mathscr{T}_q)$ (and $\theta$) can be done in one pass of the tree
(assuming the tree is already built).  Thus, not only is our bound tighter when
the cover tree imbalance is sublinear in $N$, it more closely reflects the
actual behavior of dual-tree algorithms, and the constants which it depends upon
are straightforward to calculate.

### 2015-08-14 bounds-files/imbalance.tex
i_t -> I_t, i_n -> I_n.

\begin{defn}
The {\it cover node imbalance} $i_n(\mathscr{N}_i)$ for a cover tree node
$\mathscr{N}_i$ with scale $s_i$ in the cover tree $\mathscr{T}$ is defined as
the cumulative number of missing levels between the node and its parent
$\mathscr{N}_p$ (which has scale $s_p$).  If
the node is a leaf child (that is, $s_i = -\infty$), then number of missing
levels is defined as the difference between $s_p$ and $s_{\min} - 1$ where
$s_{\min}$ is the smallest scale of a non-leaf node in $\mathscr{T}$.  If
$\mathscr{N}_i$ is the root of the tree, then the cover node imbalance is 0.
Explicitly written, this calculation is

\begin{equation}
i_n(\mathscr{N}_i) = \begin{dcases*}
  s_p - s_i - 1 & if $\mathscr{N}_i$ is not a leaf and not the root node \\
  \max(s_p - s_{\min} - 1, \; 0) & if $\mathscr{N}_i$ is a leaf \\
  0 & if $\mathscr{N}_i$ is the root node.
  \end{dcases*}
  \label{eqn_node_imbalance}
\end{equation}
\end{defn}

@@

\begin{defn}
\label{def:imbalance}
The {\it cover tree imbalance} $i_t(\mathscr{T})$ for a cover tree $\mathscr{T}$
is defined as the cumulative number of missing levels in the tree.  This can be
expressed as a function of cover node imbalances easily:

\begin{equation}
i_t(\mathscr{T}) = \sum_{\mathscr{N}_i \in \mathscr{T}} i_n(\mathscr{N}_i).
\end{equation}
\end{defn}

A perfectly balanced cover tree $\mathscr{T}_b$ with no missing levels has
imbalance $i_t(\mathscr{T}_b) = 0$ (for instance, Figure
\ref{fig:imbalance-good}).  A worst-case cover tree $\mathscr{T}_w$ which is
entirely a chain-like structure with maximum scale $s_{\max}$ and minimum scale
$s_{\min}$ will have imbalance $i_t(\mathscr{T}_w) \sim N (s_{\max} -
s_{\min})$.  Because of this chain-like structure, each level has only one node
and thus there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning
that in the worst case the imbalance is quadratic in $N$\footnote{Note that in
this situation, $c \sim N$ also.}.

>>>>

\begin{defn}
The {\it cover node imbalance} $I_n(\mathscr{N}_i)$ for a cover tree node
$\mathscr{N}_i$ with scale $s_i$ in the cover tree $\mathscr{T}$ is defined as
the cumulative number of missing levels between the node and its parent
$\mathscr{N}_p$ (which has scale $s_p$).  If
the node is a leaf child (that is, $s_i = -\infty$), then number of missing
levels is defined as the difference between $s_p$ and $s_{\min} - 1$ where
$s_{\min}$ is the smallest scale of a non-leaf node in $\mathscr{T}$.  If
$\mathscr{N}_i$ is the root of the tree, then the cover node imbalance is 0.
Explicitly written, this calculation is

\begin{equation}
I_n(\mathscr{N}_i) = \begin{dcases*}
  s_p - s_i - 1 & if $\mathscr{N}_i$ is not a leaf and not the root node \\
  \max(s_p - s_{\min} - 1, \; 0) & if $\mathscr{N}_i$ is a leaf \\
  0 & if $\mathscr{N}_i$ is the root node.
  \end{dcases*}
  \label{eqn_node_imbalance}
\end{equation}
\end{defn}

@@

\begin{defn}
\label{def:imbalance}
The {\it cover tree imbalance} $I_t(\mathscr{T})$ for a cover tree $\mathscr{T}$
is defined as the cumulative number of missing levels in the tree.  This can be
expressed as a function of cover node imbalances easily:

\begin{equation}
I_t(\mathscr{T}) = \sum_{\mathscr{N}_i \in \mathscr{T}} I_n(\mathscr{N}_i).
\end{equation}
\end{defn}

A perfectly balanced cover tree $\mathscr{T}_b$ with no missing levels has
imbalance $I_t(\mathscr{T}_b) = 0$ (for instance, Figure
\ref{fig:imbalance-good}).  A worst-case cover tree $\mathscr{T}_w$ which is
entirely a chain-like structure with maximum scale $s_{\max}$ and minimum scale
$s_{\min}$ will have imbalance $I_t(\mathscr{T}_w) \sim N (s_{\max} -
s_{\min})$.  Because of this chain-like structure, each level has only one node
and thus there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning
that in the worst case the imbalance is quadratic in $N$\footnote{Note that in
this situation, $c \sim N$ also.}.

### 2015-08-14 bounds-files/sec02-nearest-neighbor-search.tex
i_t -> I_t, i_n -> I_n.

\begin{thm}
Using cover trees, the standard cover tree pruning dual-tree traversal, and the
nearest neighbor search \texttt{BaseCase()} and \texttt{Score()} as given in
Algorithms \ref{alg:nn_base_case} and \ref{alg:nn_score}, respectively, and also
given a reference set $S_r$ with expansion constant $c_r$, and a query set $S_q$,
the running time of the algorithm is bounded by $O(c_r^4 c_{qr}^5 (N +
i_t(\mathscr{T}_q) + \theta))$ with $i_t(\mathscr{T}_q$ and $\theta$ defined as
in Definition \ref{def:imbalance} and Lemma \ref{lem:extcase3}, respectively.
\label{thm:nns}
\end{thm}

\begin{proof}
The running time of \texttt{BaseCase()} and \texttt{Score()} are clearly $O(1)$.
Due to Theorem \ref{thm:ct-runtime}, we therefore know that the runtime of the
algorithm is bounded by $O(c_r^4 |R^*| (N + i_t(\mathscr{T}_q) + \theta))$.
Thus, the only thing that remains is to bound the maximum size of the reference
set, $|R^*|$.

@@

In the monochromatic case where $S_q = S_r$, the bound is $O(c^9 (N +
i_t(\mathscr{T}))$ because $c = c_r = c_{qr}$ and $\theta = 0$.  For
well-behaved trees where $i_t(\mathscr{T}_q)$ is linear or sublinear in $N$,
this represents the current tightest worst-case runtime bound for nearest
neighbor search.

>>>>

\begin{thm}
Using cover trees, the standard cover tree pruning dual-tree traversal, and the
nearest neighbor search \texttt{BaseCase()} and \texttt{Score()} as given in
Algorithms \ref{alg:nn_base_case} and \ref{alg:nn_score}, respectively, and also
given a reference set $S_r$ with expansion constant $c_r$, and a query set $S_q$,
the running time of the algorithm is bounded by $O(c_r^4 c_{qr}^5 (N +
I_t(\mathscr{T}_q) + \theta))$ with $I_t(\mathscr{T}_q$ and $\theta$ defined as
in Definition \ref{def:imbalance} and Lemma \ref{lem:extcase3}, respectively.
\label{thm:nns}
\end{thm}

\begin{proof}
The running time of \texttt{BaseCase()} and \texttt{Score()} are clearly $O(1)$.
Due to Theorem \ref{thm:ct-runtime}, we therefore know that the runtime of the
algorithm is bounded by $O(c_r^4 |R^*| (N + I_t(\mathscr{T}_q) + \theta))$.
Thus, the only thing that remains is to bound the maximum size of the reference
set, $|R^*|$.

@@

In the monochromatic case where $S_q = S_r$, the bound is $O(c^9 (N +
I_t(\mathscr{T}))$ because $c = c_r = c_{qr}$ and $\theta = 0$.  For
well-behaved trees where $I_t(\mathscr{T}_q)$ is linear or sublinear in $N$,
this represents the current tightest worst-case runtime bound for nearest
neighbor search.

### 2015-08-14 bounds-files/sec03-approx-kde.tex
Some minor last-minute changes.
i_t -> I_t, i_n -> I_n.

\begin{thm}
Assume that $\mathcal{K}(\cdot, \cdot)$ is a kernel satisfying the assumptions
of the previous subsection.  Then, given a query set $S_q$ and a reference set
$S_r$ with expansion constant $c_r$, and using the approximate kernel density
estimation \texttt{BaseCase()} and \texttt{Score()} as given in Algorithms
\ref{alg:kde_base_case} and \ref{alg:kde_score}, respectively, with the
traversal given in Algorithm \ref{alg:cover-tree-dual}, the running time of
approximate kernel density estimation for some error parameter $\epsilon$ is
bounded by
%\begin{equation}
$O(c_r^{8 + \lceil \log_2 \zeta \rceil} (N + i_t(\mathscr{T}_q) + \theta))$
%\end{equation}
with $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$,
$i_t(\mathscr{T}_q)$ defined as in Definition \ref{def:imbalance}, and $\theta$
defined as in Lemma \ref{lem:extcase3}.

@@

\begin{proof}
It is clear that \texttt{BaseCase()} and \texttt{Score()} both take $O(1)$ time,
so Theorem \ref{thm:ct-runtime} implies the total runtime of the dual-tree
algorithm is bounded by $O(c_r^4 |R^*| (N + i_t(\mathscr{T}_q) + \theta))$.
Thus, we will bound $|R^*|$ using techniques related to those used by
\citet{ram2009}.  The bounding of $|R^*|$ is split into two sections: first,
we show that when the scale $s_r^{\max}$ is small enough, $R^*$ is empty.
Second, we bound $R^*$ when $s_r^{\max}$ is larger.

>>>>

\begin{thm}
Assume that $\mathcal{K}(\cdot, \cdot)$ is a kernel satisfying the assumptions
of the previous subsection.  Then, given a query set $S_q$ and a reference set
$S_r$ with expansion constant $c_r$, and using the approximate kernel density
estimation \texttt{BaseCase()} and \texttt{Score()} as given in Algorithms
\ref{alg:kde_base_case} and \ref{alg:kde_score}, respectively, with the
traversal given in Algorithm \ref{alg:cover-tree-dual}, the running time of
approximate kernel density estimation for some error parameter $\epsilon$ is
bounded by
%\begin{equation}
$O(c_r^{8 + \lceil \log_2 \zeta \rceil} (N + I_t(\mathscr{T}_q) + \theta))$
%\end{equation}
with $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$,
$I_t(\mathscr{T}_q)$ defined as in Definition \ref{def:imbalance}, and $\theta$
defined as in Lemma \ref{lem:extcase3}.

@@

\begin{proof}
It is clear that \texttt{BaseCase()} and \texttt{Score()} both take $O(1)$ time,
so Theorem \ref{thm:ct-runtime} implies the total runtime of the dual-tree
algorithm is bounded by $O(c_r^4 |R^*| (N + I_t(\mathscr{T}_q) + \theta))$.
Thus, we will bound $|R^*|$ using techniques related to those used by
\citet{ram2009}.  The bounding of $|R^*|$ is split into two sections: first,
we show that when the scale $s_r^{\max}$ is small enough, $R^*$ is empty.
Second, we bound $R^*$ when $s_r^{\max}$ is larger.

### 2015-08-14 bounds-files/sec04-range-search.tex
Bibliography changes, I want to stab myself, I hate this shit so much.
Some minor last-minute changes.
i_t -> I_t, i_n -> I_n.

@@

\begin{equation}
$S[p_q] = \{ p_r \in S_r : d(p_q, p_r) \in [l, u] \}$
\end{equation}

for each query point $p_q$, where $[l, u]$ is the given
range.  The range count problem is practically identical, but only the size of
the set, $|S[p_q]|$, is desired.
Our proof works for both of these algorithms
similarly, but we will focus on range search.  A \texttt{BaseCase()} and
\texttt{Score()} function are given in Algorithms \ref{alg:rs_bc} and
\ref{alg:rs_sc}, respectively \citep[a correctness proof can be found
in][]{curtin2013tree}.  The sets $N[p_q]$ (for each $p_q$) are
initialized to $\emptyset$ at the beginning of the traversal.

@@

\begin{equation}
O\left(c_r^{4} \max\left(c_r^{4 + \beta}, |S_{\max}| + C\right) (N +
i_t(\mathscr{N}_q) + \theta)
\right)
\end{equation}

@@

\begin{proof}
Both \texttt{BaseCase()} (Algorithm \ref{alg:rs_bc}) and \texttt{Score()}
(Algorithm \ref{alg:rs_sc}) take $O(1)$ time.  Therefore, using Lemma
\ref{thm:ct-runtime}, we know that the runtime of the algorithm is bounded by
$O(c_r^4 |R^*| (N + i_t(\mathscr{N}_q) + \theta))$.  As with the previous
proofs, then, our only task is to bound the maximum size of the reference set,
$|R^*|$.

@@

\begin{equation}
O(c_r^{8 + \beta} (N + i_t(\mathscr{N}_q) + \theta)).
\end{equation}
\label{cor:rs}
\end{cor}

@@

In this setting we can more easily consider the relation of the running time to
$\alpha$.  Consider $\alpha = (1 / 3)$; this yields a running time of $O(c^8 (N
+ \theta))$.  $\alpha = (1 / 7)$ yields $O(c^9 (N + i_t(\mathscr{N}_q +
\theta))$, $\alpha = (1 / 15)$ yields $O(c^{10} (N + i_t(\mathscr{N}_q) +
\theta))$, and so forth.  As $\alpha$ gets smaller, the exponent on $c$ gets
larger, and diverges as $\alpha \to 0$.

>>>>

\begin{equation}
S[p_q] = \{ p_r \in S_r : d(p_q, p_r) \in [l, u] \}
\end{equation}

\noindent for each query point $p_q$, where $[l, u]$ is the given
range.  The range count problem is practically identical, but only the size of
the set, $|S[p_q]|$, is desired.
Our proof works for both of these algorithms
similarly, but we will focus on range search.  A \texttt{BaseCase()} and
\texttt{Score()} function are given in Algorithms \ref{alg:rs_bc} and
\ref{alg:rs_sc}, respectively \citep[a correctness proof can be found
in][]{curtin2013tree}.  The sets $N[p_q]$ (for each $p_q$) are
initialized to $\emptyset$ at the beginning of the traversal.

@@

@@

\begin{equation}
O\left(c_r^{4} \max\left(c_r^{4 + \beta}, |S_{\max}| + C\right) (N +
I_t(\mathscr{N}_q) + \theta)
\right)
\end{equation}

@@

\begin{proof}
Both \texttt{BaseCase()} (Algorithm \ref{alg:rs_bc}) and \texttt{Score()}
(Algorithm \ref{alg:rs_sc}) take $O(1)$ time.  Therefore, using Lemma
\ref{thm:ct-runtime}, we know that the runtime of the algorithm is bounded by
$O(c_r^4 |R^*| (N + I_t(\mathscr{N}_q) + \theta))$.  As with the previous
proofs, then, our only task is to bound the maximum size of the reference set,
$|R^*|$.

@@

\begin{equation}
O(c_r^{8 + \beta} (N + I_t(\mathscr{N}_q) + \theta)).
\end{equation}
\label{cor:rs}
\end{cor}

@@

In this setting we can more easily consider the relation of the running time to
$\alpha$.  Consider $\alpha = (1 / 3)$; this yields a running time of $O(c^8 (N
+ \theta))$.  $\alpha = (1 / 7)$ yields $O(c^9 (N + I_t(\mathscr{N}_q +
\theta))$, $\alpha = (1 / 15)$ yields $O(c^{10} (N + I_t(\mathscr{N}_q) +
\theta))$, and so forth.  As $\alpha$ gets smaller, the exponent on $c$ gets
larger, and diverges as $\alpha \to 0$.
